
---
title: "NYPD Shooting Incidents Analysis"
author: "Hayden Stone"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: flatly
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**#Import libraries**
```{r import_libraries, echo=TRUE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(maps)
library(effects)
```

**#Load data**

#From NYC OpenData: "List of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year."

#About Page URL: https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/833y-fsy8/about_data
```{r read_in_data, echo=TRUE}
data_url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv"
nyc_data <- read_csv(data_url)
```

**#Remove Null Values For Key Variables**
```{r geo_clean, echo=TRUE}
nyc_data_filtered <- nyc_data %>% filter(!is.na(Latitude) & !is.na(Longitude))
nyc_data_filtered <- nyc_data_filtered %>%
  filter(!is.na(VIC_RACE) & VIC_RACE != "" & VIC_RACE != "NULL" & VIC_RACE != "(null)")
nyc_data_filtered <- nyc_data_filtered %>%
  filter(!is.na(PERP_RACE) & PERP_RACE != "" & PERP_RACE != "NULL"& PERP_RACE != "(null)")
nyc_data_filtered <- nyc_data_filtered %>%
  filter(!is.na(PERP_SEX) & PERP_SEX != "" & PERP_SEX != "NULL" & PERP_SEX != "(null)")
nyc_data_filtered <- nyc_data_filtered %>%
  filter(!is.na(VIC_SEX) & VIC_SEX != "" & VIC_SEX != "NULL" & VIC_SEX != "(null)")
```
### Analysis A - Analyze Victim Sex vs Perpetrator Sex
```{r analyze_sex, echo=TRUE}
#Create contingency table
counts <- table(nyc_data_filtered$VIC_SEX, nyc_data_filtered$PERP_SEX)
counts_df <- as.data.frame(counts)
counts_df <- counts_df[counts_df$Freq > 0, ]

#Make plot
ggplot(counts_df, aes(Var1, Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Victim Sex", y = "Count", fill = "Perpetrator Sex") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

### Analysis B - Analyze Race Dynamics
```{r analyze_race_bubbles, echo=TRUE}
#Create contingency table
counts <- table(nyc_data_filtered$VIC_RACE, nyc_data_filtered$PERP_RACE)
counts_df <- as.data.frame(counts)
counts_df <- counts_df[counts_df$Freq > 0, ]

#Make plot
ggplot(counts_df, aes(Var1, Var2, size = Freq)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(0, 10), 
                        breaks = c(1000, 3000, 5000, 7000), 
                        labels = c("1000", "3000", "5000", "7000"),
                        guide = guide_legend(override.aes = list(alpha = .7))) +
  labs(x = "Victim Race", y = "Perpetrator Race", size = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Make Frequency Table
vic_race_freq <- nyc_data_filtered %>%
  group_by(VIC_RACE) %>%
  summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  arrange(desc(percentage)) 
print(vic_race_freq)

```

### Analysis C - Analyze Boro vs Victim Race
```{r analyze_boro, echo=TRUE}
#Create contingency table
counts <- table(nyc_data_filtered$BORO, nyc_data_filtered$VIC_RACE)
counts_df <- as.data.frame(counts)
counts_df <- counts_df[counts_df$Freq > 0, ]

#Make plot
ggplot(counts_df, aes(Var1, Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Boro", y = "Count", fill = "Victim Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.text = element_text(size = 5),
        legend.title = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"))
```

**#Make Base Maps**
```{r get_base_maps, echo=TRUE}
# Get the base map of NYC
nyc_counties <- c("new york", "kings", "queens", "bronx", "richmond")
nyc_map <- map_data("county", region = "new york") %>%
  filter(subregion %in% nyc_counties)
```

### Analysis D - Map Black Victims
```{r make_black_victim_map, echo=TRUE}
#Filter data
nyc_data_filtered_murder <- subset(nyc_data_filtered,VIC_RACE == "BLACK")

#Make map
ggplot() +
  geom_polygon(data = nyc_map, aes(x = long, y = lat, group = group), fill = "gray80", color = "white") +
  geom_point(data = nyc_data_filtered_murder, aes(x = Longitude, y = Latitude), alpha = 1, size = .2) +
  coord_quickmap() +
  labs(title = "Black Victims in NYC", x = "Longitude", y = "Latitude", color = "Black Victim") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.text = element_text(size = 5),
        legend.title = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"))
```

### Analysis E - Map White Hispanic Victims
```{r make_white_hispanic_victim_map, echo=TRUE}

#Filter data
nyc_data_filtered_murder <- subset(nyc_data_filtered,VIC_RACE == "WHITE HISPANIC")

#Make map
ggplot() +
  geom_polygon(data = nyc_map, aes(x = long, y = lat, group = group), fill = "gray80", color = "white") +
  geom_point(data = nyc_data_filtered_murder, aes(x = Longitude, y = Latitude), alpha = 1, size = .2) +
  coord_quickmap() +
  labs(title = "White Hispanic Victims in NYC", x = "Longitude", y = "Latitude", color = "White Hispanic Victim") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.text = element_text(size = 5),
        legend.title = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"))
```

### Analysis F - Analyze "Same Race" Dynamics

```{r same_race_analysis, echo=TRUE}
#Create new SAME_RACE variable for records where VIC_RACE == PERP_RACE
nyc_data_filtered <- nyc_data_filtered %>%
  mutate(SAME_RACE = VIC_RACE == PERP_RACE)

#Calculate percentages
shares_df <- nyc_data_filtered %>%
  group_by(SAME_RACE) %>%
  summarise(
    total_shootings = n()
  ) %>%
  mutate(Share = total_shootings / sum(total_shootings) * 100,
         Share_Label = paste0(round(Share, 0), "%"))

#Make plot
ggplot(shares_df, aes(x = "", y = Share, fill = as.factor(SAME_RACE))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = Share_Label), position = position_stack(vjust = 0.5), data = shares_df) +
  labs(x = NULL, y = NULL, fill = "Same Race", title = "Share of Shootings by Same Race") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks = element_blank())
```

### Analysis G - Analyze "Same Race" Dynamics for Shootings Resulting in Murder

```{r same_race_murder_analysis, echo=TRUE}

#Logical variable transformation
nyc_data_filtered <- nyc_data_filtered %>%
  mutate(SAME_RACE = as.logical(SAME_RACE),
         STATISTICAL_MURDER_FLAG = as.logical(STATISTICAL_MURDER_FLAG))

#Calculate percentages
shares_df <- nyc_data_filtered %>%
  group_by(SAME_RACE) %>%
  summarise(
    total_shootings = n(),
    murders = sum(STATISTICAL_MURDER_FLAG == TRUE),
    non_murders = total_shootings - murders
  ) %>%
  gather(key = "Outcome", value = "Count", murders, non_murders) %>%
  mutate(Share = Count / total_shootings * 100,
         Share_Label = paste0(round(Share, 0), "%"))

#Make plot
ggplot(shares_df, aes(x = "", y = Share, fill = Outcome)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  facet_wrap(~ SAME_RACE) +
  geom_text(aes(label = Share_Label), position = position_stack(vjust = 0.5), data = shares_df) +
  labs(x = NULL, y = NULL, fill = "Outcome", title = "Share of Shootings Resulting in Murders by Same Race") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks = element_blank())
```

### Analysis H - Model Same Race as a Predictor for Shooting Resulting in a Murder

```{r create_model, echo=TRUE}

#Create model
model <- glm(STATISTICAL_MURDER_FLAG ~ SAME_RACE, data = nyc_data_filtered, family = binomial)

#Summarize model
summary(model)
```

**#Session Info**
```{r sessioninfo, echo=TRUE}
sessionInfo()
```
